---
title: "Student Performance"
author: "Martin Blasi"
date: "2023-03-15"
output: html_document
---
```{r}
#setwd("C:/MARTIN/UPF/UPF Studies/Second trimester/MSC/final project/1STUDENT/student")
library(tidyverse)
library(mombf)
mat <- read.csv2('../data/student-mat.csv')
por <- read.csv2('../data/student-por.csv')
```


#Ideas
1) Explore relationship of grades GP1-3 - interesting patterns?
+Other descriptive summaries: 
2) from MDA: perform PCA
3) Build a predictive model: Goal - A)what factors are important in student performance & B) What techniques can we use to build a good predictive model
- binary: 0/1 fail/pass (logistic)
- Categorical: splitting grades in 5 categories (count data, poisson)
- with full grades: 0-20 (also count data, poisson, negative binomial)
=>use regression techniques and find adequate specification (model comparison)
=>Use cross-validation
Optional
4) Go deeper and use Data Mining techniques like in paper: Random forest - compare with results from 3)
5) Is this even a multilevel case? (lecture 9) Consider that?


```{r}
mat.covariates = select(mat, -c("G1", "G2", "G3"))
mat.G1 = select(mat, -c("G2", "G3"))
fit1<- glm(G1 ~ ., data = mat.G1)
summary(fit1)

mat.G2 = select(mat, -c("G1", "G3"))
fit2<- glm(G2 ~ ., data = mat.G2)
summary(fit2)

```

```{r}
fit3 <- bestBIC(G3 ~ ., data = mat, family = "poisson")
summary(fit3)

fit4 <- glm(G3 ~ absences + G2, data = mat, family = "poisson")
summary(fit4)
```


