---
title: "Student Performance"
author: "Martin Blasi"
date: "2023-03-15"
output: html_document
---
```{r}
#setwd("C:/MARTIN/UPF/UPF Studies/Second trimester/MSC/final project/1STUDENT/student")
library(tidyverse)
library(mombf)
mat <- read.csv2('../data/student-mat.csv')
por <- read.csv2('../data/student-por.csv')
```

# Attributes for both student-mat.csv (Math course) and student-por.csv (Portuguese language course) datasets:
1 school - student's school (binary: "GP" - Gabriel Pereira or "MS" - Mousinho da Silveira)
2 sex - student's sex (binary: "F" - female or "M" - male)
3 age - student's age (numeric: from 15 to 22)
4 address - student's home address type (binary: "U" - urban or "R" - rural)
5 famsize - family size (binary: "LE3" - less or equal to 3 or "GT3" - greater than 3)
6 Pstatus - parent's cohabitation status (binary: "T" - living together or "A" - apart)
7 Medu - mother's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
8 Fedu - father's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
9 Mjob - mother's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
10 Fjob - father's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
11 reason - reason to choose this school (nominal: close to "home", school "reputation", "course" preference or "other")
12 guardian - student's guardian (nominal: "mother", "father" or "other")
13 traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
15 failures - number of past class failures (numeric: n if 1<=n<3, else 4)
16 schoolsup - extra educational support (binary: yes or no)
17 famsup - family educational support (binary: yes or no)
18 paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
19 activities - extra-curricular activities (binary: yes or no)
20 nursery - attended nursery school (binary: yes or no)
21 higher - wants to take higher education (binary: yes or no)
22 internet - Internet access at home (binary: yes or no)
23 romantic - with a romantic relationship (binary: yes or no)
24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25 freetime - free time after school (numeric: from 1 - very low to 5 - very high)
26 goout - going out with friends (numeric: from 1 - very low to 5 - very high)
27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
29 health - current health status (numeric: from 1 - very bad to 5 - very good)
30 absences - number of school absences (numeric: from 0 to 93)

# these grades are related with the course subject, Math or Portuguese:
31 G1 - first period grade (numeric: from 0 to 20)
31 G2 - second period grade (numeric: from 0 to 20)
32 G3 - final grade (numeric: from 0 to 20, output target)

Additional note: there are several (382) students that belong to both datasets . 
These students can be identified by searching for identical attributes
that characterize each student, as shown in the annexed R file.

#The final project 
-->should involve data analysis and R programming, using the methods seen in class. 
The project must feature a regression problem with multiple covariates, and must have at least one of the following features: 
• The outcome is a binary or count variable 
• The model includes non-linear covariate effects 
• The data and analysis method features a multi-level structure 

Make sure to discuss model checking, interactions and non-linearities, and whether it was necessary to use computational methods to obtain confidence intervals / P-values. 

You should turn in: 
(1) You should turn in a report in pdf, with no R code (see below for its format). If it contains animations, the format can be in html but its printed size (when saved via File-> Print to file) cannot exceed the page limit (9 pages)
(2) The markdown file with the R code used to produce the analyses 
(3) A short dashboard or interactive document in html format displaying your main results. It should be 1 pages with ≤ 3 tabs containing only a few key figures / tables. Also turn in the markdown file with the code used to produce the document 

• Final report due: March 23rd, 2023 

#Ideas
1) Explore relationship of grades GP1-3 - interesting patterns?
+Other descriptive summaries: 
2) from MDA: perform PCA
3) Build a predictive model: Goal - A)what factors are important in student performance & B) What techniques can we use to build a good predictive model
- binary: 0/1 fail/pass (logistic)
- Categorical: splitting grades in 5 categories (count data, poisson)
- with full grades: 0-20 (also count data, poisson, negative binomial)
=>use regression techniques and find adequate specification (model comparison)
=>Use cross-validation
Optional
4) Go deeper and use Data Mining techniques like in paper: Random forest - compare with results from 3)
5) Is this even a multilevel case? (lecture 9) Consider that?


```{r}
mat.covariates = select(mat, -c("G1", "G2", "G3"))
mat.G1 = select(mat, -c("G2", "G3"))
fit1<- glm(G1 ~ ., data = mat.G1)
summary(fit1)

mat.G2 = select(mat, -c("G1", "G3"))
fit2<- glm(G2 ~ ., data = mat.G2)
summary(fit2)

```

```{r}
fit3 <- bestBIC(G3 ~ ., data = mat, family = "poisson")
summary(fit3)

fit4 <- glm(G3 ~ absences + G2, data = mat, family = "poisson")
summary(fit4)
```


